How fluid-mechanical erosion creates anisotropic porous media

### 2022-06-17 docs/PhysicsPaper/StokesErosion.tex
NM

Now consider measuring the transverse permeability $k_{22}$ using Darcy's law \eqref{eqn:Darcy}. Taking the vertical component of \eqref{eqn:Darcy} yields
\begin{equation}
\label{q1}
q_2(x,y) = -k_{22} \pderiv{p}{y}(x,y)
\end{equation}
but this form is not useful if no vertical pressure gradient is imposed, as is the case in our erosion simulations. It is important to recognize that the permeability is a property of the {\em medium}, not the imposed flow. Hence, for a frozen configuration of bodies, it is permissible to alter the imposed flow for the purpose of measuring $k_{22}$ (this altered flow is completely separate from the simulation of the erosion process that generates the configurations). Thus, instead of a horizontal flow in the far-field, we seek to impose a vertical one $\UU = (0,U)$.
If this were to be done directly, the outer geometry would need to be rotated by 90 degrees about the configuration of bodies inside (which does not rotate). In practice, it is simpler to keep the outer geometry fixed and instead rotate the inner configuration of bodies, then simply apply the method from Section \ref{LongPerm} to measure $k_{22}$.

With both the longitudinal and transverse components of permeability calculated, we define the anisotropy of the medium as the ratio
\begin{equation}
\anis = k_{11} / k_{22}
\end{equation}
Note that the random configuration of circles used to initialize the erosion simulations will have an anisotropy nearly equal to one. As this configuration erodes, it would be expected to permit flow in the longitudinal direction more easily than in the transverse direction, yielding $\anis > 1$.

There are two mechanistic explanations for how erosion can create this anisotropy. First, the shear stresses could carve each individual body into a more slender form, thus creating anisotropy at the level of individual grains. Second, by removing material in a non-uniform way, could preferentially disintegrate certain bodies before others, leaving an overall configuration that exhibits anisotropy on the large scale, for example one with horizontally aligned channels. We will refer to these two possible mechanisms as {\em shape} anisotropy and {\em configurational} anisotropy respectively. As an extreme example, a tightly-packed horizontal row of circular bodies would exhibit configurational anisotropy but no shape anisotropy. Meanwhile, an array of highly eccentric ellipses, all oriented horizontally but positioned randomly, would exhibit high shape anisotropy and low configurational anisotropy.

@@

We note that channelization is related to both shape and configurational anisotropy....

It is possible to devise a test to isolate these two types of anisotropy.
Introduce circular bodies with same centers of mass...
Thus, the configurational anisotropy is defined as
\begin{equation}
\anis_C = k_1^{(c)} / k_2^{(c)}
\end{equation}

The total anisotropy is the product of the shape and configurational
\begin{equation}
\anis = \anis_S \anis_C
\end{equation}

>>>>

Now consider measuring the transverse permeability $k_{22}$ using Darcy's law \eqref{eqn:Darcy}. Taking the vertical component of \eqref{eqn:Darcy} yields
\begin{equation}
\label{q1}
q_2(x,y) = -k_{22} \pderiv{p}{y}(x,y)
\end{equation}
This form, however, is not useful if no vertical pressure gradient is imposed, as is the case in our erosion simulations. It is important to recognize that the permeability is a property of the {\em medium}, not the imposed flow. Hence, for a frozen configuration of bodies, it is permissible to alter the imposed flow for the purpose of measuring $k_{22}$ (this altered flow is completely separate from the simulation of the erosion process that generates the configurations). Thus, instead of a horizontal flow in the far-field, we seek to impose a vertical one $\UU = (0,U)$.
If this were to be done directly, the outer geometry would need to rotate by 90 degrees about the fixed configuration of bodies. In practice, it is simpler to keep the outer geometry fixed and rotate the inner configuration of bodies, then simply apply the method from Section \ref{LongPerm} to measure $k_{22}$.

With both the longitudinal and transverse components of permeability computed, we define the anisotropy of the medium as the ratio between the two:
\begin{equation}
\anis = k_{11} / k_{22}
\end{equation}
Note that the random configuration of circles used to initialize the erosion simulations will have an anisotropy nearly equal to one. As this configuration erodes, it would be expected to permit flow in the longitudinal direction more easily than in the transverse direction, yielding $\anis > 1$.

There are two mechanistic explanations for how erosion can create medium anisotropy. First, the shear stresses could carve each individual body into a more slender form, thus creating anisotropy at the level of individual grains. Second, the shear stressed could preferentially remove certain bodies before others, thus creating large-scale anisotropy. For example, a body positioned in a high-throughput channel might disintegrate relatively quickly, thus opening the channel and creating higher overall anisotropy. We will refer to these two possible mechanisms as {\em shape} anisotropy and {\em configurational} anisotropy respectively. As an extreme example, a tightly-packed horizontal row of circular bodies would exhibit high configurational anisotropy but no shape anisotropy. Meanwhile, an array of highly eccentric ellipses, all oriented horizontally but positioned randomly, would exhibit high shape anisotropy and low configurational anisotropy. We note that channelization is likely associated with {\em both} types of anisotropy. Certainly the overall configuration must support a channel, and, secondly, the geometry of the individual bodies that outline the channel could control its structure to some degree.

	Fortunately, it is possible to devise a test to isolate these two types of anisotropy from one another and therefore determine how much each contributes to the total anisotropy. In particular, for a fixed configuration of partially-eroded bodies, we replace each body with a circle having the same area and center of mass. Since the configurations resulted from the erosion of initially circular bodies, such a replacement does not lead to any overlap between bodies. We then measure the two permeabilities, $k_{11}^{(c)}$ and $ k_{22}^{(c)}$, of this configuration of circular bodies. Since all of the individual shapes are identical and isotropic (i.e. circles), the resulting anisotropy is entirely due to the relative positions of the bodies, thus allowing us to define the {\em configurational} anisotropy as $\anis_C = k_{11}^{(c)} / k_{22}^{(c)}$. The total anisotropy, meanwhile, is the product of the shape and configurational anisotropy, $\anis = \anis_S \anis_C$, the latter of which we have measured. We can therefore deduce the shape anisotropy through $\anis_S = \anis/\anis_C$.

@@
